Theory for taxonomic bias and differential abundance

### 2021-10-27 theory/derivations-regression.Rmd
Split original theory out from the manuscript
The original theoretical calculations, which have been put in the
appendix, have largely become redundant and/or superfluous to the
manuscript. In order to improve repo organization and streamline the
rendered manuscript, this commit moves these files to a new theory/
folder and removes them from the rendered manuscript. The theory files
remain structured as a separate Bookdown article.

```{theorem, label = "regression-linearity", name = "Linearity of regression coefficients", echo = TRUE}
Consider a scalar response variable with $J$ observations, which I represent by the vector $y$, that equals a sum of $K$ component response variables $y^{(k)}$ each scaled by a non-zero factor $c^{(k)}$, 
\begin{align}
  (\#eq:y-sum)
  y = \sum_{k = 1}^K c^{(k)} y^{(k)}.
\end{align}
(We may need to assume that the $y^{(k)}$ are linearly independent.)
Let $X$ be a $J$-by-$p$ matrix of covariates for the $J$ observations, with linearly independent columns.
Let $\hat \gamma$ denote the least-squares estimate of the coefficient matrix $\gamma$ in the linear regression
\begin{align}
  y = X \gamma + \epsilon.
\end{align}
Similarly, let $\hat \gamma^{(k)}$ denote the least-squares estimates for the $K$ linear regressions
\begin{align}
  y^{(k)} = X \gamma^{(k)} + \epsilon^{(k)}.
\end{align}
The least-squares coefficient estimates for $y$ are given by the sum of those of the $y^{(k)}$,
\begin{align}
  \hat \gamma = \sum_{k=1}^K c^{(k)} \hat \gamma^{(k)}.
\end{align}
```

```{proof, echo = TRUE}
The Moore-Penrose pseudoinverse of $X$ is $X^+ = (X^T X)^{-1} X^T$.
The least squares estimates are given by multiplying the matrix $X^+$ by the corresponding response vector ([Wikipedia](https://en.wikipedia.org/wiki/Ordinary_least_squares#Matrix/vector_formulation)), so that $\hat \gamma = X^+ y$ and $\hat \gamma^{(k)} = X^+ y^{(k)}$ (for $k = 1, \dots, K$).
The result then follows from the stipulation \@ref(eq:y-sum) and the linearity of matrix multiplication,
\begin{align}
  \hat \gamma 
  = X^+ y 
  = X^+ \sum_{k = 1}^K c^{(k)} y^{(k)}
  = \sum_{k = 1}^K c^{(k)} X^+ y^{(k)}
  = \sum_{k = 1}^K c^{(k)} \hat \gamma^{(k)}.
\end{align}
```

@@

```{theorem, label = "regression-error", echo = TRUE, name = "Error in regression coefficients"}
Suppose that $y$ is the response variable we wish to understand, $z$ is our imperfect measurement of $y$, and $d = z - y$ the difference between the two, so that $z = y + d$.
Let each be a vector of length $J$ describing a set of $J$ observations.
Let $X$ be a $J$-by-$p$ matrix of covariates for the $J$ observations, with linearly independent columns.
Consider the linear regression equations for $y$, $d$, and $z$,
\begin{align}
  y &= X \gamma^{(y)} + \epsilon^{(y)} \\
  z &= X \gamma^{(z)} + \epsilon^{(z)} \\
  d &= X \gamma^{(d)} + \epsilon^{(d)}.
\end{align}
The relationship between the least-squares estimates mirrors that between the variables themselves,
\begin{align}
  (\#eq:regression-coefficient-error)
  \hat \gamma^{(z)} = \hat \gamma^{(y)} + \hat \gamma^{(d)}.
\end{align}
```

```{proof, echo = TRUE}
The result follows directly from $z =  y + d$ and \@ref(thm:regression-linearity).
```

```{corollary, label = "regression-error-simple", echo = TRUE}
For the special case of simple linear regression,
\begin{align}
  y &= \gamma_0^{(y)} + \gamma_1^{(y)} x + \epsilon^{(y)} \\
  z &= \gamma_0^{(z)} + \gamma_1^{(z)} x + \epsilon^{(z)} \\
  d &= \gamma_0^{(d)} + \gamma_1^{(d)} x + \epsilon^{(d)},
\end{align}
the estimates for the intercept and slope coefficients satisfy
\begin{align}
  \hat \gamma_0^{(z)} &= \hat \gamma_0^{(y)} + \hat \gamma_0^{(d)} \\
  \hat \gamma_1^{(z)} &= \hat \gamma_1^{(y)} + \hat \gamma_1^{(d)}.
\end{align}
```

@@

```{proposition, echo = TRUE}
Consider the linear equations
\begin{align}
  (\#eq:least-squares-formulation)
  \log a_i    &= \alpha_0 + \alpha_1 x \\
  \log \bar B &= \beta_0 + \beta_1 x   \\
  \log m_i    &= \gamma_0 + \gamma_1 x,
\end{align}
and let $\hat \alpha_0$, etc. denote the least-squares estimates for the regression coefficients supposing perfect information (i.e., that the $a_i$ and $B_i$ are known).
The coefficients for $\log m_i$ are related to those for $\log a_i$ and $\log \bar B$ through the equations
\begin{align}
  (\#eq:least-squares-estimates)
  \hat \gamma_0 &= \hat \alpha_0 + \log B_i - \hat \beta_0 \\
  \hat \gamma_1 &= \hat \alpha_1 - \hat \beta_1.
\end{align}
```

```{proof, echo = TRUE}
```

>>>>

::: {.theorem #regression-linearity name="Linearity of regression coefficients"}
Consider a scalar response variable with $J$ observations, which I represent by the vector $y$, that equals a sum of $K$ component response variables $y^{(k)}$ each scaled by a non-zero factor $c^{(k)}$, 
\begin{align}
  (\#eq:y-sum)
  y = \sum_{k = 1}^K c^{(k)} y^{(k)}.
\end{align}
(We may need to assume that the $y^{(k)}$ are linearly independent.)
Let $X$ be a $J$-by-$p$ matrix of covariates for the $J$ observations, with linearly independent columns.
Let $\hat \gamma$ denote the least-squares estimate of the coefficient matrix $\gamma$ in the linear regression
\begin{align}
  y = X \gamma + \epsilon.
\end{align}
Similarly, let $\hat \gamma^{(k)}$ denote the least-squares estimates for the $K$ linear regressions
\begin{align}
  y^{(k)} = X \gamma^{(k)} + \epsilon^{(k)}.
\end{align}
The least-squares coefficient estimates for $y$ are given by the sum of those of the $y^{(k)}$,
\begin{align}
  \hat \gamma = \sum_{k=1}^K c^{(k)} \hat \gamma^{(k)}.
\end{align}
:::

::: {.proof}
The Moore-Penrose pseudoinverse of $X$ is $X^+ = (X^T X)^{-1} X^T$.
The least squares estimates are given by multiplying the matrix $X^+$ by the corresponding response vector ([Wikipedia](https://en.wikipedia.org/wiki/Ordinary_least_squares#Matrix/vector_formulation)), so that $\hat \gamma = X^+ y$ and $\hat \gamma^{(k)} = X^+ y^{(k)}$ (for $k = 1, \dots, K$).
The result then follows from the stipulation \@ref(eq:y-sum) and the linearity of matrix multiplication,
\begin{align}
  \hat \gamma 
  = X^+ y 
  = X^+ \sum_{k = 1}^K c^{(k)} y^{(k)}
  = \sum_{k = 1}^K c^{(k)} X^+ y^{(k)}
  = \sum_{k = 1}^K c^{(k)} \hat \gamma^{(k)}.
\end{align}
:::

@@

::: {.theorem #regression-error name="Error in regression coefficients"}
Suppose that $y$ is the response variable we wish to understand, $z$ is our imperfect measurement of $y$, and $d = z - y$ the difference between the two, so that $z = y + d$.
Let each be a vector of length $J$ describing a set of $J$ observations.
Let $X$ be a $J$-by-$p$ matrix of covariates for the $J$ observations, with linearly independent columns.
Consider the linear regression equations for $y$, $d$, and $z$,
\begin{align}
  y &= X \gamma^{(y)} + \epsilon^{(y)} \\
  z &= X \gamma^{(z)} + \epsilon^{(z)} \\
  d &= X \gamma^{(d)} + \epsilon^{(d)}.
\end{align}
The relationship between the least-squares estimates mirrors that between the variables themselves,
\begin{align}
  (\#eq:regression-coefficient-error)
  \hat \gamma^{(z)} = \hat \gamma^{(y)} + \hat \gamma^{(d)}.
\end{align}
:::

::: {.proof}
The result follows directly from $z =  y + d$ and \@ref(thm:regression-linearity).
:::

::: {.corollary #regression-error-simple}
For the special case of simple linear regression,
\begin{align}
  y &= \gamma_0^{(y)} + \gamma_1^{(y)} x + \epsilon^{(y)} \\
  z &= \gamma_0^{(z)} + \gamma_1^{(z)} x + \epsilon^{(z)} \\
  d &= \gamma_0^{(d)} + \gamma_1^{(d)} x + \epsilon^{(d)},
\end{align}
the estimates for the intercept and slope coefficients satisfy
\begin{align}
  \hat \gamma_0^{(z)} &= \hat \gamma_0^{(y)} + \hat \gamma_0^{(d)} \\
  \hat \gamma_1^{(z)} &= \hat \gamma_1^{(y)} + \hat \gamma_1^{(d)}.
\end{align}
:::

@@

::: {.proposition}
Consider the linear equations
\begin{align}
  (\#eq:least-squares-formulation)
  \log a_i    &= \alpha_0 + \alpha_1 x \\
  \log \bar B &= \beta_0 + \beta_1 x   \\
  \log m_i    &= \gamma_0 + \gamma_1 x,
\end{align}
and let $\hat \alpha_0$, etc. denote the least-squares estimates for the regression coefficients supposing perfect information (i.e., that the $a_i$ and $B_i$ are known).
The coefficients for $\log m_i$ are related to those for $\log a_i$ and $\log \bar B$ through the equations
\begin{align}
  (\#eq:least-squares-estimates)
  \hat \gamma_0 &= \hat \alpha_0 + \log B_i - \hat \beta_0 \\
  \hat \gamma_1 &= \hat \alpha_1 - \hat \beta_1.
\end{align}
:::
